Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2019-03-04 manuscript/homophily_manuscript.Rmd

Writing papers, networking, and collaboration are all instrumental to research productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], and dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; studies listed in Table S1 of @Holman_2018]. For example, studies have concluded that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], gender differences in confidence or self-esteem [@bleidorn2016age], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] or travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better outputs on collaborative tasks than less diverse teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. For reasons such as these, multiple studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, most or all such studies imply that men co-publish with men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing has often been termed 'gender homophily'.

However, we believe that earlier studies of gender homophily were hindered by a largely unacknowledged statistical issue that we will refer to as the Wahlund effect (\autoref{wahlund_plot}), by analogy with the conceptually similar Wahlund effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting the number of same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues even if no gender preferences exist, or if the true preference is for opposite-gendered colleagues ('gender heterophily'). For example, a sample of literature containing a mixture of bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers usually collaborate within their own discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

@@

There was no indication that journals publishing on a wide range of topics have higher $\alpha'$ values than more specialised journals, due to the Wahlund effect. For example, the journal category 'Multidisciplinary' -- which includes journals like _PLOS ONE_, _Nature_, _Science_, and _PNAS_ -- did not have markedly elevated $\alpha'$ (\autoref{alpha_histograms}). This result suggests that our estimates of homophily, and estimates from some of the earlier studies listed in the Introduction, are probably not inflated by the presence of disparate research topics (with variable author gender ratios) being published within individual journals.

@@

An important limitation of our study is that we cannot reliably determine the cause(s) of the observed excess of same-gender coauthorships. As well as the obvious interpretation -- conscious or unconscious selection of same-gender collaborators by men, by women, or both -- our results could be partly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these uncontrolled artefacts to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals relative to specialised journals; restricting the data by country yielded similar estimates of $\alpha'$; and we used a simple model to show that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data suggest that it is likely that some researchers preferentially select same-gendered collaborators, although it is difficult to ascertain what proportion of people show such a preference, or how much the strength of the preference varies between individuals. We also note that even in a world in which collaboration was completely random with respect to gender, a high proportion of individual researchers would have entirely same-gendered collaborators by chance alone (especially in gender-biased disciplines); thus, individuals who only have same-gendered co-authors are not necessarily doing anything differently from people with gender-balanced co-authors.

>>>>

Writing papers, networking, and collaboration are all instrumental to research productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], and dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; studies listed in Table S1 of @Holman_2018]. For example, studies have concluded that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], gender differences in confidence or self-esteem [@bleidorn2016age], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] or travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better outputs on collaborative tasks than less diverse teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. For reasons such as these, multiple studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, most or all such studies imply that men co-publish with men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This non-random assortment is often termed 'gender homophily'.

We believe that most or all earlier studies of gender homophily were hindered by a largely unacknowledged statistical issue that we will refer to as the Wahlund effect (\autoref{wahlund_plot}), by analogy with the conceptually similar Wahlund effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting the number of same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues even if no gender preferences exist, or if the true preference is for opposite-gendered colleagues ('gender heterophily'). For example, a sample of literature containing a mixture of bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers usually collaborate within their own discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

@@

There was no indication that journals publishing on a wide range of topics have higher $\alpha'$ values than more specialised journals, due to the Wahlund effect. For example, the journal category 'Multidisciplinary' -- which includes journals like _PLOS ONE_, _Nature_, _Science_, and _PNAS_ -- did not have markedly elevated $\alpha'$ (\autoref{alpha_histograms}). This result suggests that our estimates of homophily, and estimates from some of the earlier studies listed in the Introduction, are probably not markedly inflated by the presence of disparate research topics (with variable author gender ratios) being published within individual journals.

@@

An important limitation of our study is that we cannot reliably determine the cause(s) of the observed excess of same-gender coauthorships. As well as the obvious interpretation -- conscious or unconscious selection of same-gender collaborators by men, by women, or both -- our results could be partly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these uncontrolled artefacts to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals relative to specialised journals; restricting the data by country yielded similar estimates of $\alpha'$; and we used a simple model to show that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data suggest that it is likely that some researchers preferentially select same-gendered collaborators, although it is difficult to ascertain what proportion of people show such a preference, or how much the strength of the preference varies between individuals. We also note that even in a world in which everyone selected their collaborators at random with respect to gender, a high proportion of individual researchers would have entirely same-gendered collaborators by chance alone (especially in gender-biased disciplines); thus, individuals who only have same-gendered co-authors are not necessarily doing anything differently from people with gender-balanced co-authors.
